Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-17 _tex/paper2.tex
Built site for gh-pages

@@

For instance, If researchers cannot conduct the randomized experiments
described in Section~\ref{sec-appendix-A} and must instead rely on
observational data, they can still identify the ACE as long as a
variable \(X\), such as the socio-economic status of the school,
satisfies the backdoor criterion. Under these circumstances, researchers
firt identify the \emph{conditional average causal effect} (CACE,
Equation~\ref{eq-CACE})

@@

From the CACE, they identify the ACE from associational quantities as in
Equation~\ref{eq-mACE}. Here, \(E_{X}[\cdot]\) represents the marginal
expected value over \(X\) \citep{Morgan_et_al_2014}.

\begin{equation}\phantomsection\label{eq-mACE}{
\begin{aligned}
  \tau & = E[Y_{i} \:|\:do(T_{i}=1)] - E[Y_{i} \:|\:do(T_{i}=2)] \\
  & = E_{X}\left[ E[Y_{i} \:|\:T_{i}=1, X] - E[Y_{i} \:|\:T_{i}=2, X] \right]
\end{aligned}
}\end{equation}

@@

\subsection{SCMs and DAGs}\label{sec-appendix-C}

@@

Researchers can then use these building blocks to represent the scenario
outlined in Section~\ref{sec-appendix-B}. SCM \ref{fig-scm_example1} and
DAG \ref{fig-mdag_example1} depict the plausible causal structure for
this example. In this context, the variable \(X\) (socio-economic status
of the school) acts as a confounder in the relationship between the
teaching method \(T\) and the outcome \(Y\). The figures display
multiple descendant relationships such as \(X \rightarrow T\),
\(X \rightarrow Y\), and \(T \rightarrow Y\). They also highlight
unconnected node pairs, evident from the relationships
\(e_{T} \:\bot\:e_{X}\), \(e_{T} \:\bot\:e_{Y}\), and
\(e_{X} \:\bot\:e_{Y}\). Additional, the figures show one fork,
\(X \rightarrow \{T, Y\}\), and two colliders:
\(\{X, e_{T}\} \rightarrow T\) and \(\{X, T, e_{Y}\} \rightarrow Y\).

@@

\caption{\label{fig-example}Plausible causal structure the scenario
outlined in Section~\ref{sec-appendix-B}.}

@@

\subsection{The probabilistic implications of SCMs and
DAGs}\label{sec-appendix-D}

@@

The five fundamental building blocks described in
Section~\ref{sec-appendix-C} clearly illustrate which (in)dependencies
can SMCs and DAGs predict. For instance, applying the BNF to the causal
structure shown in the SCM \ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1}
supports the theoretical assumption that the observed variables
\(X_{1}\) and \(X_{3}\) are unconditionally independent or d-separated
(\(X_{1} \:\bot\:X_{3}\)). This conclusion follows from expressing their
joint probability distribution as
\(P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})\) \citep[pp.~24]{Neal_2020}.
Conversely, when \(X_{3}\) is unconditionally dependent on \(X_{1}\)
(\(X_{1} \:\not\bot\:X_{3}\)), as depicted in the SCM \ref{fig-scm_bb2}
and DAG \ref{fig-mdag_bb2}, the BNF express their joint probability
distribution as \(P(X_{1}, X_{3}) = P(X_{3} \:|\:X_{1}) P(X_{1})\).
Notably, these descriptions demonstrate the clear correspondence between
the structural equations illustrated in Section~\ref{sec-appendix-C} and
the CPDs.

@@

Researchers can revisit the scenario in Section~\ref{sec-appendix-B}
using these insights. By applying the BNF to the SCM
\ref{fig-scm_example1} and DAG \ref{fig-mdag_example1a} structures, they
can represent the joint probability of the observed variables as
\(P(Y, T, X) = P(Y \:|\:T, X) P(T \:|\:X) P(X)\). Assuming the
\emph{direct effect} of \(T\) on \(Y\) (\(T \rightarrow Y\), black path)
is absent, researchers can infer from the joint probability that \(Y\)
is not unconditionally independent of \(T\) \((Y \:\bot\:T)\). This
dependency arises because a non-causal path exists between \(T\) and
\(Y\) through \(X\) (\(T \leftarrow X \rightarrow Y\), gray path),
indicating that \(X\) is a confounder. However, researchers can observe
that conditioning on \(X\), depicted as a square in DAG
\ref{fig-mdag_example1a}, results in \(Y \:\bot\:T \:|\:X\), eliminating
the confounding effect and allowing the identification of the direct
effect \(T \rightarrow Y\) if it exists.

@@

\[
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  T & := f_{T}(X,e_{T}) \\
  Y & := f_{Y}(t,X,e_{Y}) \\
  e_{T} & \:\bot\:e_{X} \\
  e_{T} & \:\bot\:e_{Y} \\
  e_{X} & \:\bot\:e_{Y}
\end{aligned}
\]

@@

@@

\subsection{From probability to causality}\label{sec-appendix-E}

In addition to the assumptions required to encode probabilistic
information, SCMs and DAGs rely on the modularity assumption to
represent interventions \(P(Y_{i} \:|\:do(T_{i}=t))\) (see
Section~\ref{sec-appendix-B}). Defining these interventions enables
researchers to identify the ACE from associational quantities through
statistical conditioning, effectively disentangling causation from mere
association.

The modularity assumption underpins the concepts of Truncated
Factorization and manipulated graphs, which are essential for
representing interventions within SCMs and DAGs. The \emph{modularity
assumption} posits that intervening on a node modifies only the causal
mechanism of that node, leaving others unchanged
\citep[pp.~34]{Neal_2020}. Building on this principle, the
\emph{Truncated Factorization} (TF) simulates physical interventions by
removing specific functions from the conceptual model and substituting
them with constants, preserving the rest of the structure unchanged
{[}Pearl\_2010{]}. The probabilistic implications of this factorization
are formalized in Equation~\ref{eq-trunc_factor}, where \(S\) denotes
the subset of variables \(X_{p}\) directly influenced by the
intervention. Complementing the TF, \emph{manipulated graphs} are DAGs
where specific edges have been removed to reflect the effects of
interventions \citep[pp.~34]{Neal_2020}.

@@

Using the TF, researchers can define the \emph{backdoor adjustment}, a
method for identifying causal effects. This adjustment specifies that if
a variable \(X_{p} \in S\) serves as a \emph{sufficient adjustment set}
for the effect of \(X_{a}\) on \(X_{b}\), then the ACE can be identified
using Equation~\ref{eq-backdoor} (see Section~\ref{sec-appendix-B}).

@@

@@

@@

\caption{\label{fig-example}Plausible causal structure the scenario
outlined in Section~\ref{sec-appendix-B}.}

@@

\subsection{The estimation process}\label{sec-appendix-F}

Researchers can use Bayesian inference methods to estimate the ACE. The
approach begins by defining two probability distributions: the
likelihood of the data, \(P(X_{1}, X_{2}, \dots, X_{P} \:|\:\theta)\),
and the prior distribution, \(P(\theta)\) \citep{Everitt_et_al_2010},
where \(X_{P}\) represents a random variable, and \(\theta\) represents
a one-dimensional parameter space for simplicity. After observing
empirical data, researchers can update the priors to posterior
distributions using Bayes' rule \citep{Jeffreys_1998}:

@@

Nevertheless, researchers can manage the complexity of the likelihood by
assuming that variables exhibit specific local (in)dependencies. These
assumptions enhance model tractability and facilitate the estimation
process. As described in Section~\ref{sec-appendix-D}, researchers can
use SCMs and DAGs to formalize these local (in)dependencies, which leads
to the BNF of the likelihood (Equation~\ref{eq-like_BNF}). From this
point, any probabilistic programming language can model this simpler
structure and generate the parameter's posterior distribution from its
priors using Equation~\ref{eq-bayes_rule}.

>>>>

\subsubsection{The general idea}\label{sec-appendix-B1}

@@

For instance, If researchers cannot conduct the randomized experiments
described in Section~\ref{sec-appendix-A} and must instead rely on
observational data, they can still identify the ACE as long as a
variable \(X\), such as the socio-economic status of the school,
satisfies the backdoor criterion. Under these circumstances, researchers
first identify the \emph{conditional average causal effect} (CACE,
Equation~\ref{eq-CACE})

@@

From the CACE, they identify the ACE from associational quantities as in
Equation~\ref{eq-mACE1}. This identification process is commonly known
as the \emph{backdoor adjustment}. Here, \(E_{X}[\cdot]\) represents the
marginal expected value over \(X\) \citep{Morgan_et_al_2014}.

\begin{equation}\phantomsection\label{eq-mACE1}{
\begin{aligned}
  \tau & = E[Y_{i} \:|\:do(T_{i}=1)] - E[Y_{i} \:|\:do(T_{i}=2)] \\
  & = E_{X}\left[ E[Y_{i} \:|\:T_{i}=1, X] - E[Y_{i} \:|\:T_{i}=2, X] \right]
\end{aligned}
}\end{equation}

@@

\subsubsection{Diving into the specifics}\label{sec-appendix-B2}

\paragraph{SCMs and DAGs}\label{sec-appendix-B21}

@@

Researchers can then use these building blocks to represent the scenario
outlined in Section~\ref{sec-appendix-B1}. SCM \ref{fig-scm_example1}
and DAG \ref{fig-mdag_example1} depict the plausible causal structure
for this example. In this context, the variable \(X\) (socio-economic
status of the school) acts as a confounder in the relationship between
the teaching method \(T\) and the outcome \(Y\). The figures display
multiple descendant relationships such as \(X \rightarrow T\),
\(X \rightarrow Y\), and \(T \rightarrow Y\). They also highlight
unconnected node pairs, evident from the relationships
\(e_{T} \:\bot\:e_{X}\), \(e_{T} \:\bot\:e_{Y}\), and
\(e_{X} \:\bot\:e_{Y}\). Additional, the figures show one fork,
\(X \rightarrow \{T, Y\}\), and two colliders:
\(\{X, e_{T}\} \rightarrow T\) and \(\{X, T, e_{Y}\} \rightarrow Y\).

@@

\caption{\label{fig-example}Plausible causal structure the scenario
outlined in Section~\ref{sec-appendix-B1}.}

@@

\paragraph{The probabilistic implications of SCMs and
DAGs}\label{sec-appendix-B22}

@@

The five fundamental building blocks described in
Section~\ref{sec-appendix-B21} clearly illustrate which (in)dependencies
can SMCs and DAGs predict. For instance, applying the BNF to the causal
structure shown in the SCM \ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1}
supports the theoretical assumption that the observed variables
\(X_{1}\) and \(X_{3}\) are unconditionally independent or d-separated
(\(X_{1} \:\bot\:X_{3}\)). This conclusion follows from expressing their
joint probability distribution as
\(P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})\) \citep[pp.~24]{Neal_2020}.
Conversely, when \(X_{3}\) is unconditionally dependent on \(X_{1}\)
(\(X_{1} \:\not\bot\:X_{3}\)), as depicted in the SCM \ref{fig-scm_bb2}
and DAG \ref{fig-mdag_bb2}, the BNF express their joint probability
distribution as \(P(X_{1}, X_{3}) = P(X_{3} \:|\:X_{1}) P(X_{1})\).
Notably, these descriptions demonstrate the clear correspondence between
the structural equations illustrated in Section~\ref{sec-appendix-B21}
and the CPDs.

@@

Researchers can revisit the scenario in Section~\ref{sec-appendix-B1}
using these insights. By applying the BNF to the SCM
\ref{fig-scm_example1a} and DAG \ref{fig-mdag_example1a} structures,
they can represent the joint probability of the observed variables as
\(P(Y, T, X) =\) \(P(Y \:|\:T, X) P(T \:|\:X) P(X)\). From this
expression, researchers can infer that the outcome \(Y\) is not
unconditionally independent of the teaching method \(T\)
\((Y \:\bot\:T)\). This dependency arises from two key structures: a
direct causal path from the teaching method \(T\) to the outcome \(Y\),
represented by the two-connected-nodes structure \(T \rightarrow Y\)
(black path), and a confounding non-causal path from the teaching method
\(T\) to the outcome \(Y\) through the socio-economic status of the
school \(X\), represented by the fork structure
\(T \leftarrow X \rightarrow Y\) (gray path).

These probabilistic insights also indicate that narrowing the focus to a
subset of the population defined by the socio-economic status of the
school \(X\), achieved by conditioning on \(X\) and depicted by a square
in DAG \ref{fig-mdag_example1a}, effectively blocks all confounding
non-causal path introduced by the fork structure. Nevertheless, since
\(Y\) remains not d-separated from \(T\) even after conditioning
\((Y \:\not\bot\:T \:|\:X)\), this lack of d-separation can only be
attributed to the direct causal effect \(T \rightarrow Y\).

@@

\[
\begin{aligned}
  X & := x \\
  T & := f_{T}(x,e_{T}) \\
  Y & := f_{Y}(T,x,e_{Y}) \\
  e_{T} & \:\bot\:e_{X} \\
  e_{T} & \:\bot\:e_{Y} \\
  e_{X} & \:\bot\:e_{Y}
\end{aligned}
\]

@@

\subcaption{\label{fig-scm_example1a}SCM}

@@

\paragraph{From probability to causality}\label{sec-appendix-B23}

The structural approach to causal inference translate the probabilistic
insights from Section~\ref{sec-appendix-B23} into actionable strategies
for identifying the ACE from associational quantities. The approach
achieve this by relying on the modularity assumption and defining
interventions of the form \(P(Y_{i} \:|\:do(T_{i}=t))\) (see
Section~\ref{sec-appendix-B1}). The \emph{modularity assumption} posits
that intervening on a node alters only the causal mechanism of that
node, leaving others unchanged \citep[pp.~34]{Neal_2020}.

The modularity assumption underpins the concepts of manipulated graphs
and Truncated Factorization, which are essential for representing
interventions within SCMs and DAGs. \emph{Manipulated graphs} simulate
physical interventions by removing specific edges from a DAG, while
preserving the remaining structure unchanged \citep[pp.~34]{Neal_2020}.
In parallel, \emph{Truncated Factorization} (TF) achieves a similar
simulation by removing specific functions from the conceptual model and
replacing them with constants, while keeping the rest of the structure
unchanged {[}Pearl\_2010{]}. The probabilistic implications of this
factorization are formalized in Equation~\ref{eq-trunc_factor}, where
\(S\) represents the subset of variables \(X_{p}\) directly influenced
by the intervention.

@@

Using the TF, researchers can define the \emph{backdoor adjustment} to
identify the ACE. This adjustment states that if a variable
\(X_{p} \in S\) serves as a \emph{sufficient adjustment set} for the
effect of \(X_{a}\) on \(X_{b}\), then the ACE can be identified using
Equation~\ref{eq-backdoor}. The sufficient adjustment set (potentially
empty) must block all non-causal paths between \(X_{a}\) and \(X_{b}\)
without introducing new paths. If such a set exists, then \(X_{a}\) and
\(X_{b}\) are \emph{d-separated} by \(X_{p}\)
(\(X_{a} \:\bot\:X_{b} \:|\:X_{p}\)) \citep{Pearl_2009}, and \(X_{p}\)
satisfies the \emph{backdoor criterion} \citep[pp.~37]{Neal_2020}.

@@

With these new insights, researchers revisiting the scenario in
Section~\ref{sec-appendix-B22} can infer that the socio-economic status
of the school, \(X\), satisfies the backdoor criterion, assuming the
causal structure depicted by the SCM \ref{fig-scm_example1a} and DAG
\ref{fig-mdag_example1a} is correct. This means that \(X\) serves as a
sufficient adjustment set, as it effectively blocks all confounding
non-causal paths introduced by the fork structure. In this context,
researchers can apply the \emph{backdoor adjustment} to identify the ACE
of \(T\) on \(Y\). They achieve this by first identifying the CACE of
\(T\) on \(Y\) by conditioning on \(X\), and then marginalizing this
effect over \(X\) to obtain the ACE, as expressed in
Equation~\ref{eq-mACE2} (see Section~\ref{sec-appendix-B}). Notably, for
the purpose of identification, the conditioned DAG
\ref{fig-mdag_example1a} is equivalent to the manipulated DAG
\ref{fig-mdag_example1b}, because \(X\) satisfies the backdoor
criterion.

\begin{equation}\phantomsection\label{eq-mACE2}{
\begin{aligned}
  \tau & = E[Y_{i} \:|\:do(T_{i}=1)] - E[Y_{i} \:|\:do(T_{i}=2)] \\
  & = E_{X}\left[ E[Y_{i} \:|\:T_{i}=1, X] - E[Y_{i} \:|\:T_{i}=2, X] \right]
\end{aligned}
}\end{equation}

@@

\subcaption{\label{fig-scm_example1b}SCM}

@@

\caption{\label{fig-example}Plausible causal structure the scenario
outlined in Section~\ref{sec-appendix-B22}.}

@@

\subsection{The estimation process}\label{sec-appendix-C}

Ultimately, researchers can use Bayesian inference methods to estimate
the ACE. The approach begins by defining two probability distributions:
the likelihood of the data,
\(P(X_{1}, X_{2}, \dots, X_{P} \:|\:\theta)\), and the prior
distribution, \(P(\theta)\) \citep{Everitt_et_al_2010}, where \(X_{P}\)
represents a random variable, and \(\theta\) represents a
one-dimensional parameter space for simplicity. After observing
empirical data, researchers can update the priors to posterior
distributions using Bayes' rule \citep{Jeffreys_1998}:

@@

Nevertheless, researchers can manage the complexity of the likelihood by
assuming that variables exhibit specific local (in)dependencies. These
assumptions enhance model tractability and facilitate the estimation
process. As described in Section~\ref{sec-appendix-B22}, researchers can
use SCMs and DAGs to formalize these local (in)dependencies, which leads
to the BNF of the likelihood (Equation~\ref{eq-like_BNF}). From this
point, any probabilistic programming language can model this simpler
structure and generate the parameter's posterior distribution from its
priors using Equation~\ref{eq-bayes_rule}.
